---
title: "mcmc-sampling"
output: html_document
date: "2025-05-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Metropolis Hastings demo

-type of MCMC method to obtain random (correlated) samples, often for high-dimensional distribution

-markov chain=sample t+1 depends on sample t, markov process asymptotically reaches stationary distribution; monte carlo=random

-for conjugate distribution,posterior and prior=same distribution. compute parameters for posterior.

-for posterior, when it is not analytically solvable, construct markov chain to do monte carlo approximation

-other MCMC: Gibbs (conditional), HMC (efficient steps)

Note: Code was generated with the help of GPT-4o.

Source: <https://github.com/twiecki/WhileMyMCMCGentlySamples/blob/master/content/downloads/notebooks/MCMC-sampling-for-dummies.ipynb>

```{r cars}
summary(cars)
```

```{r}
# Load necessary libraries
library(ggplot2)

# Function to plot prior distribution, histogram, and mu lines
plot_distribution <- function(data, mu_current, mu_proposal, mu_prior, std_prior, mu_true=NULL) {
  # Create a sequence of x values for the prior distribution
  x <- seq(min(data) - 3, max(data) + 3, length.out = 100)
  
  # Calculate the prior PDF
  prior_pdf <- dnorm(x, mean = mu_prior, sd = std_prior)
  
  # Create a data frame for plotting
  plot_data <- data.frame(x = x, prior_pdf = prior_pdf)
  
  # Create the histogram of the data
  hist_data <- data.frame(data = data)
  
  # Plot
  result<-ggplot() +
    geom_histogram(data = hist_data, aes(x = data, y = ..density..), 
                   binwidth = 0.5, fill = "grey", alpha = 0.5) +                # data distribution
    geom_line(data = plot_data, aes(x = x, y = prior_pdf), color = "black") +   # prior distribution
    geom_vline(xintercept = mu_current, color = "blue", linetype = "dashed", size = 1) +    # current mu
    geom_vline(xintercept = mu_proposal, color = "green", linetype = "dashed", size = 1) + # proposed mu
    annotate("text", x = mu_current, y = max(prior_pdf) * 0.85, label = "Current mu", color = "blue", vjust = -1) +
    annotate("text", x = mu_proposal, y = max(prior_pdf) * 0.85, label = "Proposed mu", color = "green", vjust = -1) +
    geom_segment(aes(x = mu_current, xend = mu_proposal, y = 0, yend = 0), 
                 arrow = arrow(type = "closed", length = unit(0.2, "inches")), 
                 color = "black") +
    labs(title = "Prior Distribution and Data Histogram",
         x = "Value",
         y = "Density") +
    theme_minimal()
  if (!is.null(mu_true)) {
    result <- result + 
      geom_vline(xintercept = mu_true, color = "red", linetype = "dashed", size = 1) +
      annotate("text", x = mu_true, y = max(prior_pdf) * 0.85, label = "True mu", color = "red", vjust = -1)
  }
  result
}

```

```{r}
sampler <- function(data, samples = 4, mu_init = .5, proposal_width = .5, plot = FALSE, mu_prior_mu = 0, mu_prior_sd = 1., mu_true=NULL) {
  mu_current <- mu_init
  posterior <- list()
  
  for (i in seq(from = 1, to = samples)) {
    # proposed mean, proposal width=jump width
    mu_proposal <- rnorm(1, mean=mu_current, sd=proposal_width)
    
    # compute data likelihoods (assume iid)
    likelihood_current <- prod(dnorm(data, mean = mu_current, sd = 1))
    likelihood_proposal <- prod(dnorm(data, mean = mu_proposal, sd = 1))
    
    # compute prior probability of current and proposed mu (from prior distribution)
    prior_current <- dnorm(mu_current, mean = mu_prior_mu, sd = mu_prior_sd)
    prior_proposal <- dnorm(mu_proposal, mean = mu_prior_mu, sd = mu_prior_sd)
    
    p_current <- likelihood_current * prior_current
    p_proposal <- likelihood_proposal * prior_proposal
    
    # Accept proposal? (ratio cancels p(x))
    p_accept <- p_proposal / p_current
    
    # Usually would include prior probability, which we neglect here for simplicity
    accept <- runif(1) < p_accept
    
    if (plot && i%%10==0) {
      plot_result <- plot_distribution(data, mu_current, mu_proposal, mu_prior_mu, mu_prior_sd, mu_true)
      print(plot_result)
    }

    if (accept) {
      # Update position
      mu_current <- mu_proposal
    } else {
      mu_current <- mu_current
    }
    
    posterior[[length(posterior) + 1]] <- mu_current
  }
  
  posterior
}
```

```{r echo=FALSE}
# initialize variables
data <- cars$speed
mu_init <- mean(data) + runif(1)*3 #choose a good mu, otherwise it may get nan's and get stuck 
proposal_width = .2
mu_prior_mu = mean(data) # usually we wont have good estimate
mu_prior_sd = 1.

result <- sampler(data = data, samples = 100, mu_init = mu_init, proposal_width = proposal_width, plot = TRUE, mu_prior_mu = mu_prior_mu, mu_prior_sd = mu_prior_sd, mu_true = mean(data))
```
